Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course

### 2024-08-04 paper/paper.qmd
First round

---
title: "[Result, but poetic]: Does LLM Use Improve Data Science Education?"
subtitle: "Evidence from a Canadian Undergraduate Statistics Course"
author: 
  - Rohan Alexander
  - Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/llms-achievement."
date: today
date-format: long
abstract: "An abstract should eventually go here!!"
format: pdf
number-sections: true
table-of-contents: true
bibliography: references.bib
nocite: | #references.bib entries at end, regardless of intext citation
  @* 
---

```{r setup, include=FALSE}
## Workspace Setup ##

library(tidyverse)
library(knitr)
library(kableExtra)
library(lintr)
library(tidyr)
library(tidytext)
library(textdata)
library(quanteda)
library(wordcloud)
library(reshape2)
library(topicmodels)

# Import data
survey_data <- read_csv("../data/analysis_data/clean_STA302_postcourse_survey_w24.csv", show_col_types = FALSE)
```

@@

The wide release of ChatGPT and other Large Language Models (LLMs) has rapidly transformed education at all levels and disciplines, creating excitement and apprehension among educators and students alike. They offer unprecedented support for student learning in tasks ranging from writing assistance to support in complex problem-solving, with the potential to enhance academic performance and student outcomes. At the same time, their use raises important questions about academic integrity, the development of critical thinking skills (particular in undergraduates), as well as questions about their implications for effective learning overall.

The potential for LLMs to positively affect students' academic performance in data science can be clearly inferred from already-demonstrated effects on adjacent professional fields. The tasks involved in a data science workflow can be generally broken down into two key competencies. The first is programming, which is done when cleaning, analyzing, visualizing data using languages like R or Python. The second is writing, which is primarily done when communicating results. There is experimental evidence from adjacent professional fields that LLMs can improve productivity in both of these competencies.

@Peng_2023 show the potential impact of LLMs on students' computer programming skills, through their uncovering of the significantly positive impacts of GitHub Copilot (an LLM-powered programming assistant) on the productivity of computer programmers. In an experiment involving 95 freelance programmers, they found that that programmers with access to Copilot completed a standardized programming task 55.8% faster than the control group. Crucially, it was found that programmers with less experience saw the greatest improvements in productivity. Given that programming is a core component of data science practice, this suggests that these benefits could reasonably carry over to students in data science courses as well.

@DellAcqua_2023, while primarily about management consulting tasks, provides evidence that LLMs can significantly improve writing productivity. In a field experiment involving Consultants recruited from the Boston Consulting Group (BCG), they found that the use of GPT-4 in the experimental group led to a 25% increase in delivery speed of business tasks (most involving some writing), as well as a 40% increase in human-rated performance on those tasks. Similar to computer programming, these productivity increases were most pronounced for those with below average performance, with their output increasing by 43%. Though not all tasks done using AI saw the same productivity improvements, the authors expressed particular optimism for LLMs' potential to generally expedite menial knowledge-work tasks. These included tasks such as generating new ideas and creating persuasive and informative writing - both examples directly apply to data science education through their analogues of coming up with project ideas and writing more engaging reports. As a whole, they show that LLMs could equally augment the quality of student writing and course projects in data science as well.

Clearly, the implications of these case studies appear promising. However, literature from educators both adjacent to and distant from statistics and data science suggest much more varied perspectives on how LLMs can impact student performance.

@Valenzuela_2024 argue that LLMs lead to a loss of serendipity (which leads to less original work) and de-skilling (primarily with respect to programming ability), among other consequences. These particular outcomes could negatively affect students' effective learning of data science. On the other hand, @Ellis_Slade_2023 take a more optimistic perspective, taking the popular stance of comparing ChatGPT to previously controversial learning technologies like calculators. They argue that LLMs are just another technology that will impact Statistics and Data Science education like the calculator did. However, they take a broader perspective on the issue and more specific inquiry could be done in how effective LLMs are at improving student outcomes within data science education in particular.

In a similar attitude, @Tu_2024 acknowledge that for students, LLMs can streamline many parts of a data science workflow - with that in mind, they suggest that data scientists in training should shift their self-perspectives from primarily being an analyst to primarily being a manager responsible for strategic oversight of the analysis. Crucially they emphasize that in both education and practice, LLMs and human intelligence should play complementary roles.

The simultaneous caution and interest the literature expressed toward using LLMs in educational settings was further corroborated by empirical studies on K-12 and university students as well.

At the high school level, @Lazar_2023 conducted a informal survey of secondary school teachers and students on their opinions of ChatGPT, they found that while LLMs could help spark creativity, provide academic support when teachers were unavailable, and model certain types of writing well, teachers were also cognizant of LLMs potential to limit students' learning in certain ways through overreliance. Beyond academic integrity concerns, teachers had similar concerns to @Valenzuela_2024 about de-skilling and an overall loss of agency in writing and critical thinking.

@Cahill_2024 surveyed undergraduate Political Science students on their attitudes toward and usage of AI tools. They found that the use of ChatGPT (among other machine-learning-powered software) was widespread. However, they also found that many students lacked the confidence in using AI for academic purposes - in particular, only 11% 'strongly agreed' that they know how to use AI to improve their writing. Like educators,  students have nuanced views on appropriate AI use. In particular, respondents found that using it to writing whole papers as inappropriate, while using it for basic tasks like general assistance, writing feedback and basic data visualization was perceived more appropriately. Interestingly, first-generation college students were found to be more likely to use AI in their work, particularly in writing papers and helping with assignments. Their findings suggest that LLMs and other AI tools could be an equalizer for disadvantaged students. Rhough only political science students were surveyed, statistics and data science students could reasonably have general attitudes that are similar.

As we have seen, the existing inquiry by educators has explored the general qualitative usage and student perceptions on these tools. Though informative, a key question still remains: how can students' academic performance be affected, precisely, through allowing them to use LLMs in classwork?

This paper aims to fill this gap by quantitatively investigating the relationship between students' grades and measures of student LLM usage and their attitudes toward LLMs in general, using evidence from a third-year undergraduate statistics course at the University of Toronto. Through examining how students interact with and perceive LLM tools and how these variables translate into student outcomes, the effects of LLM integration in data science education can be more precisely determined.

The remainder of this paper is structured as follows: @sec-data visualizes and analyzes survey data and coursework from students; @sec-model models the unstructured data to approximate a relationship between grades and usage/attitudes; @sec-results describes and analyzes the model's results; @sec-discussion discusses the implications of the findings for data science education and future research and practice in this rapidly evolving field.

@@

## Background and Collection

To investigate students' usage and attitudes toward LLM use and how they related to their academic performance, a dataset containing their usage/attitudes, coursework, and academic performance was constructed. Data was sampled from the cohort of students taking STA302 - Methods of Data Analysis I, taught by one of the investigators in the Winter 2024 semester at the University of Toronto. By virtue of pre-requisites needed, this restricted the data collected to be only on upper-year undergraduate students.

Data was collected from students through an optional end-of-course survey. Whether or not they consented to their data being used in this investigation, all respondents received a +1% increase in their final course grade for their participation. All consenting responses were then cross-referenced to their course grade, as well as the GitHub account they used to complete their course research papers. The responses were finally anonymized by removing any personal references to the students themselves.

## Cleaning and Variables of Interest

All the data was cleaned using R [@citeR] and it's tidyverse [@tidyverse], janitor [@janitor] and stringr [@stringr], then tested for issues using the testthat package [@testthat]. This led to a final dataset containing 121 unique responses to 18  survey questions.

```{r fig-boilerplate}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: (CHANGE) Boilerplate figure caption 
#| fig-align: center
# survey_data |>
#   ggplot(program_counts, aes(x = "", y = n, fill = program_of_study)) +
#   #put geom type here (put before compilation) +
#   labs(title = "Title!",
#        x = "X label",
#        y = "Y label") +
#   theme_minimal() +
#   scale_fill_discrete(name="Legend Title") +
#   theme(plot.title.position = "plot",
#         plot.title = element_text(hjust = 0.5)
#         )
```

```{r tbl-boilerplate}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: (CHANGE) Boilerplate table caption 
#| fig-align: center
#Put code below this line
```

## General Information

First, general information about the sample of students was derived first, through visualizing the basic data about them. This was done using the ggplot2 [@ggplot2] package.

```{r fig-response-time-hist}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Distribution of Survey Response Times
#| fig-align: center

# Trim values to histogram

plot_data <- survey_data |>
  filter(mins_complete < 1000)

plot_data |>
  ggplot(aes(x = mins_complete)) +
  geom_histogram(binwidth = 2, fill = "lightblue", color = "black") +
  labs(
    title = "Distribution of Survey Response Times",
    x = "Time (Mins)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5)
  )
```
@fig-response-time-hist shows the distribution of response times to the end-of-course survey. Respondents who consented generally took between 0-40 mins to answer all of the questions, with most response times ranging between 2-10 minutes. There was one consenting respondent who took 4308 minutes, whose observation was dropped from the visualization. Overall, however, this suggests that students generally took a meaningful amount of time and thought to engage with the survey's questions.

```{r fig-gpa-hist}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Distribution of Student GPA
#| fig-align: center

# Histogram of GPA

survey_data |>
  ggplot(aes(x = gpa)) +
  geom_histogram(binwidth = 0.1, fill = "lightblue", color = "black") +
  labs(
    title = "Distribution of Student GPA",
    x = "GPA",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.6)
  )
```

@fig-gpa-hist shows a wide distribution of student GPAs, with the majority clustering around a B (3.0/4.0) average. One factor possibly affecting the range is the course being a required course for programs in the Statistics, Mathematics and Computer Science departments. Though programs in these departments generally attract high-achieving students, some may still struggle. Another factor is that these GPAs were self-reported, which introduces some ambiguity - students may have provided either their cumulative GPA or their most recent term's GPA, or could have misreported it entirely.  Despite these two factors, the diversity in reported grades suggests the course attracted students across a broad spectrum of academic performance levels.

```{r tbl-year-dist}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Count of Students by Year
#| fig-align: center

# Change PEY to 4th year
survey_data <- survey_data |>
  mutate(
    year = case_when(
      year == "PEY" ~ "4th",
      year == "5th" ~ "5th+",
      year == "6th" ~ "5th+",
      TRUE ~ year
    )
  )

year_count <- survey_data |>
  count(year) |>
  pivot_wider(names_from = year, values_from = n, values_fill = list(n = 0))

# Display the original wide format table
kable(year_count)
```

We can also see in from @tbl-year-dist that students from a range of years took the course. That said, the majority of students were in their 3rd or 4th year of study. This makes sense, given the course's prerequisite of general statistics, which is a two-semester sequence typically done in second year statistics. This also explains the lack of any first year students in the course as well.

```{r fig-programs-pie}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Student Distribution by Program of Study
#| fig-align: center

# Pie chart of program of study

program_counts <- survey_data |>
  count(program_of_study)

ggplot(program_counts, aes(x = "", y = n, fill = program_of_study)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(
    title = "Distribution of Students by Program of Study",
    subtitle = "Visualizing the proportion of students across different programs",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  scale_fill_discrete(name = "Program of Study") +
  theme(
    axis.text.x = element_blank(), # Remove x-axis text
    axis.title.x = element_blank(), # Remove x-axis title
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  geom_text(aes(label = n),
    position = position_stack(vjust = 0.5)
  )
```

Programs of study [did something; waiting on major/minor data], as we can see in @fig-programs-pie...

## Student Attitudes

```{r fig-ai-familiarity}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Student Familiarity with AI
#| fig-align: center

table_data <- survey_data |>
  select(ai_familiarity) |>
  mutate(
    ai_familiarity = case_when(
      ai_familiarity == "Limited to some liberal arts, such as history courses" ~ "Somewhat familiar",
      TRUE ~ ai_familiarity
    )
  ) |>
  count(ai_familiarity)

# Custom colors
familiarity_colors <- c(
  "Not familiar" = "tomato",
  "Somewhat familiar" = "lightblue",
  "Very familiar" = "lightgreen"
)

table_data |>
  ggplot(aes(x = "", y = n, fill = ai_familiarity)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  labs(
    title = "Student Familiarity with AI",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  scale_fill_manual(name = "Familiarity Level", values = familiarity_colors) +
  theme(
    axis.text.x = element_blank(), # Remove x-axis text
    axis.title.x = element_blank(), # Remove x-axis title
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  geom_text(aes(label = n),
    position = position_stack(vjust = 0.5)
  )
```

Just as @Cahill_2024 found, the vast majority of students taking the course were at least "somewhat familiar" with ChatGPT and other LLMs. This overrepresentation of familiarity also primarily driven by the fact that the course is primarily targeted for students in Statistics, Computer Science, Economics, Mathematics and other quantitative disciplines, possibly leading to a sample that is skewed to being more interested in computing.

Students were then asked about their attitudes and self-perception on writing and coding.

```{r fig-writing-multibar-1}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Student Self-Perceptions on Writing, Part 1
#| fig-align: center

# Create a multi-bar plot for writing perceptions grouped by response

# Select first three statements
writing_perceptions <- survey_data |>
  select(
    writing_is_easy_for_me,
    i_like_to_write,
    i_believe_it_is_important_to_be_a_good_writer
  )

writing_perceptions_long <- writing_perceptions |>
  pivot_longer(cols = everything(), names_to = "Statements", values_to = "Response") |>
  mutate(
    Statements = case_when(
      Statements == "writing_is_easy_for_me" ~ "\"Writing is easy for me\"",
      Statements == "i_like_to_write" ~ "\"I like to write\"",
      Statements == "i_believe_it_is_important_to_be_a_good_writer" ~ "\"I believe it is important\nto be a good writer\"",
      TRUE ~ Statements
    ),
    Response = case_when(
      Response == "This is a lot like me" ~ "A lot like me",
      Response == "This is very different to me" ~ "Very different from me",
      Response == "This somewhat describes me" ~ "Somewhat describes me",
    )
  ) |>
  filter(!is.na(Response))

# Custom colors
response_colors <- c(
  "A lot like me" = "lightgreen",
  "Somewhat describes me" = "lightblue",
  "Very different from me" = "tomato"
)

# Taken from: https://stackoverflow.com/questions/52297978/decrease-overal-legend-size-elements-and-text
addSmallLegend <- function(myPlot, pointSize = 0.5, textSize = 3, spaceLegend = 0.1) {
  myPlot +
    guides(
      shape = guide_legend(override.aes = list(size = pointSize)),
      color = guide_legend(override.aes = list(size = pointSize)),
      fill = guide_legend(title = "Response", override.aes = list(size = pointSize))
    ) +
    theme(
      legend.title = element_text(size = textSize),
      legend.text = element_text(size = textSize),
      legend.key.size = unit(spaceLegend, "lines")
    )
}

# Make the plot
multi_bar <- writing_perceptions_long |>
  ggplot(aes(x = Response, fill = Response)) +
  geom_bar(position = "dodge", stat = "count") +
  facet_wrap(~Statements) +
  labs(
    title = "Student Self-Perceptions on Writing",
    subtitle = "Responses to statements about writing ease, enjoyment, and importance",
    x = "Response",
    y = "Number of Students"
  ) +
  scale_fill_manual(values = response_colors, name = "Response") + # Using custom colors
  theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.x = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

multi_bar <- addSmallLegend(multi_bar, pointSize = 0.5, textSize = 9, spaceLegend = 0.8)

multi_bar
```

```{r fig-writing-multibar-2}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Student Self-Perceptions on Writing, Part 2
#| fig-align: center

# Create a multi-bar plot for writing perceptions grouped by response

# Select last three statements
writing_perceptions <- survey_data |>
  select(
    when_i_edit_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_writing,
    i_am_confident_in_my_overall_writing_ability
  )

writing_perceptions_long <- writing_perceptions |>
  pivot_longer(cols = everything(), names_to = "Statements", values_to = "Response") |>
  mutate(
    Statements = case_when(
      Statements == "when_i_edit_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I edit, it is easy to\ncatch my own mistakes\"",
      Statements == "i_feel_confident_sharing_my_writing" ~ "\"I feel confident\nsharing my writing\"",
      Statements == "i_am_confident_in_my_overall_writing_ability" ~ "\"I am confident in\nmy own writing ability\"",
      TRUE ~ Statements
    ),
    Response = case_when(
      Response == "This is a lot like me" ~ "A lot like me",
      Response == "This is very different to me" ~ "Very different from me",
      Response == "This somewhat describes me" ~ "Somewhat describes me",
    )
  ) |>
  filter(!is.na(Response))

# Custom colors
response_colors <- c(
  "A lot like me" = "lightgreen",
  "Somewhat describes me" = "lightblue",
  "Very different from me" = "tomato"
)

>>>>

---
title: "Self-reported LLM usage and results on a data science project: Evidence from a Canadian undergraduate data science course"
author: 
  - Rohan Alexander
  - Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/llms-achievement. We thank Nathalie Moon for developing and sharing her survey questions. This research is currently under review by the University of Toronto's University Research Ethics Board. Comments can be sent to: rohan.alexander@utoronto.ca."
date: today
date-format: long
abstract: "To help understand the effect of Large Language Models (LLMs) on data science practice we examine the extent to which LLM usage is correlated with the mark that a student gets on a final paper in a classroom data science setting. We find no clear relationship between more extensive LLM usage and the student's mark. Despite the classroom setting used for evaluation, the particular activity of interest reflects the work done by professional data scientists. Our finding suggests the need for more extensive work evaluating how LLMs can be integrated into the data science workflow in a way that provides value."
format: pdf
number-sections: true
table-of-contents: false
bibliography: references.bib
---

```{r setup}
#| include: false

library(knitr)
library(quanteda)
library(readxl)
library(reshape2)
library(tidytext)
library(tidyverse)

# Old data (REMOVE ONCE THE TEXT ANALYSIS BIT IS UPDATED)
survey_data <- read_csv("../data/analysis_data/clean_STA302_postcourse_survey_w24.csv", show_col_types = FALSE)

# Read in data
all_data <- read_excel(here::here("data/raw_data/merged_data.xlsx"), sheet = "Sheet1") |>
  janitor::clean_names()
```

@@

There are many aspects of trustworthy data science, including education, culture, and workflow to name a few. One aspect of special importance is the tools that we use to do data science: the computers, programming languages, settings, and environments. The wide release of user-friendly Large Language Models (LLMs), especially OpenAI's ChatGPT, is a rare instance of a new tool being made available.

Like all new tools, LLMs have created both excitement and apprehension. ChatGPT's public release on November 30, 2022, brought LLMs into the mainstream conversation. By now many people, especially educators and students, have some experience with LLMs, in both personal and professional contexts.

In the context of teaching statistics and data science LLMs could be useful for many tasks. For instance, there is considerable interest in the potential of chatbots and personalized tutors **[ADD CITE]**. ChatGPT has also been the catalyst for many interesting and important conversations around academic integrity **[ADD CITE]**, the development of critical thinking skills **[ADD CITE]**, and what effective learning looks like **[ADD CITE]**.

In this paper we are interested in better understanding LLMs as a tool for producing trustworthy data science. We study how they were used by students in an upper-year undergraduate data science course and whether students who used LLMs tended to have higher scores than those who did not.

The tasks involved in a trustworthy data science workflow can be generally broken down into a number of key competencies [@Adhikari2021Interleaving; @Gibbs2021Building]. The potential for LLMs to positively affect students' academic performance in data science can be clearly inferred from already-demonstrated effects in a number of those competencies on adjacent professional fields. The first is programming, which is done when cleaning, analyzing, visualizing data often using languages like R or Python. The second is writing, which is primarily done when communicating results.

In terms of general programming, @Peng_2023 found a positive impact of GitHub Copilot (an LLM-powered programming assistant) on productivity. Specifically, in an experiment involving 95 freelance programmers, they found that that programmers with access to GitHub Copilot completed a standardized programming task 56% faster than the control group. Programmers with less experience saw the greatest improvements in productivity.

@DellAcqua_2023 focus on management consulting tasks and provide evidence that LLMs can improve writing productivity. In a field experiment involving consultants from Boston Consulting Group they found that the use of OpenAI's GPT-4 led to a 25% increase in delivery speed of business tasks (most involving some writing), as well as a 40% increase in human-rated performance on those tasks. Similar to computer programming, these productivity increases were most pronounced for those with below average performance, with their output increasing by 43%.

On the other hand, @Valenzuela_2024 argue that LLMs lead to a loss of serendipity (which leads to less original work) and de-skilling (primarily with respect to programming ability), among other consequences. These outcomes could negatively affect students' effective learning of data science.

@Ellis_Slade_2023 take a more optimistic perspective and argue that LLMs are just another technology that will impact statistics and data science education. Similarly, @Tu_2024 acknowledge that LLMs can streamline many parts of a data science workflow. With that in mind, they suggest that data-scientists-in-training should shift their self-perspectives from primarily being an analyst to primarily being a manager responsible for strategic oversight of the analysis.

At the high school level @Lazar_2023 conducted a informal survey of secondary school teachers and students on their opinions of ChatGPT, and found that while LLMs could: help creativity, provide academic support when teachers were unavailable, and model certain types of writing well; teachers were also aware of LLMs potential to limit students' learning in certain ways through over-reliance. Beyond academic integrity concerns, teachers had similar concerns to @Valenzuela_2024 about de-skilling and an overall loss of agency in writing and critical thinking.

@Cahill_2024 surveyed undergraduate political science students on their attitudes toward and usage of AI tools. They found that the use of ChatGPT was widespread. However, they also found that many students lacked confidence in using AI for academic purposes. In particular, only 11% 'strongly agreed' that they know how to use AI to improve their writing. Students had nuanced views on appropriate AI use. In particular, respondents found that using it to write whole papers as inappropriate, while using it for basic tasks like general assistance, writing feedback and basic data visualization was perceived more appropriately.

To understand the current state of LLMs as a tool for trustworthy data science, this paper focuses on the association between student academic performance and their LLM usage. Specifically, we examine the relationship between students' grades and self-reported measures of student LLM usage, as well as student attitudes toward LLMs in general. This is based on final papers and a survey, conducted in a third-year undergraduate data science course at the University of Toronto. By examining how students interact with and perceive LLMs as tools, and how these variables translate into student outcomes, the effects of LLM integration in data science can be more precisely determined and recommendations made for future development.

The remainder of this paper is structured as follows: @sec-data visualizes and analyzes survey data and coursework from students. @sec-model specifies a model used to investigate the relationship. @sec-results describes and analyzes the model's results. @sec-discussion discusses the implications of the findings for data science education and future research and practice at the intersection of LLMs and trustworthy data science.

@@

## Background

To investigate students' usage and attitudes towards LLMs and how they related to their academic performance, a dataset containing their usage/attitudes, coursework, and academic performance was constructed. This was based on three components:

1. an optional survey; 
2. self-reported LLM usage; and 
3. student marks on the final paper.

All data are from the cohort of students taking STA302 "Methods of Data Analysis I" in the Winter 2024 semester at the University of Toronto. This course had 275 students initially enrolled which, reflecting a normal rate of attrition for undergraduate statistics courses at the University of Toronto, reduced to 154 students by the end of the semester. Assessment was heavily based on three papers submitted over the course of the 12 week semester.

The student marks that we analyze are based only on the final paper, which is done individually. By this stage, uninterested students have typically dropped the course, and students are familiar with course expectations. A typical paper submission is 10-20 pages, and requires students to conduct original research to answer a research question of interest to them. It reflects the skills typically used by a professional data scientist. Students are expected to develop a research question of interest to them, identify or collect data to answer the question, conduct statistical analysis, and write a short paper. Examples of final papers (shared with consent) include: @hannahyu; @emilysu; and @bennyrochwerg.

By the time they are working on their final paper, students have submitted and received feedback on two previous papers with similar requirements and rubrics to that of the final paper. Each paper has the same basic structure and expectations. Before the final paper is due students have received feedback on all their previous work in the class (including their past papers) and there is an, optional, two-day period of peer review.

By virtue of pre-requisites of this course the typical student is an upper-year undergraduate. Coding and writing are major parts of the course. Students are welcome to use R or Python, but the majority code in R because that is the programming language currently mostly taught in pre-requisite courses. All writing must be in English. The aim of the course for students to create a public portfolio of work they can use to apply for jobs.

Throughout the semester students were encouraged to use LLMs. Formal instruction was provided twice during the semester, with a masterclass taught by a computer science faculty member on the ethics of using LLMs (see @Horton2024 for details), and another masterclass taught by a TA on writing with LLMs.

Data was collected from students through an optional end-of-course survey. [Appendix -@sec-survey-details] provides all the questions asked in the survey. Whether or not they consented to their data being used, all respondents received a 1% increase in their final course grade for their participation. Consenting responses were then matched to their final paper mark, as well as the GitHub repository for their final paper. The responses were anonymized by removing any personal references to the students themselves including names, emails, student numbers, and GitHub links.

All data cleaning and analysis was done using the `R` statistical programming language [@citeR], and especially uses the `tidyverse` [@tidyverse], `janitor` [@janitor], `reshape2` [@reshape], and `readxl` [@readxl] libraries.

## Survey data

There were 146 responses to the survey. Of these, 119 respondents provided authorization for their data to be collected and used. Four of those respondents submitted the survey twice, and after removing their second response, 115 responses remained. Of those, 15 respondents did not include a statement on LLM usage in the README of the GitHub repository of their final paper, leaving 100 responses that were of use and were merged based on student name.

All but one respondent completed the survey within 45 minutes (@fig-response-time-and-gpa-1). That one respondent took more than 4,000 minutes to complete the survey, suggesting they took a break while filling it out. That respondent is included in the analysis dataset, but without that respondent, the average time to complete the survey was 11 minutes, and the standard deviation was 8 minutes.

```{r fig-response-time-and-gpa}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Distribution of respondents' survey response times and GPA
#| fig-subcap: ["Distribution of survey response times","\"What is your GPA?\""]
#| layout-ncol: 1

all_data |>
  mutate(mins_complete = completion_time - start_time) |> 
  filter(mins_complete < 1000) |> # Remove one student who took 4,000 minutes to complete
  # summarize(mean = mean(mins_complete),
            # standard_dev = sd(mins_complete))
  ggplot(aes(x = mins_complete)) +
  geom_histogram(binwidth = 1) +
  labs(
    x = "Time to complete survey (minutes)",
    y = "Frequency"
  ) +
  theme_minimal()

all_data |>
  # summarize(mean = mean(what_is_your_gpa, na.rm = TRUE),
  #           standard_dev = sd(what_is_your_gpa, na.rm = TRUE))
  ggplot(aes(x = what_is_your_gpa)) +
  geom_histogram(binwidth = 0.1) +
  labs(
    x = "Self-reported GPA",
    y = "Frequency"
  ) +
  theme_minimal()
```

There is a wide distribution of self-reported GPAs (@fig-response-time-and-gpa-2). Seven of the 100 respondents included in the analysis dataset did not report their GPA. The majority of responses cluster around a B (3.0/4.0), and the average is 3.06 with a standard deviation of 0.55. One factor that may affect the range is that the course is required for programs in the Statistics, Mathematics and Computer Science Departments. Self-reported GPAs introduce the possibility of reporting bias. For instance, respondents may have provided their cumulative GPA, their most recent term's GPA, or could have misreported it entirely.

Students from a range of years took the course, however the majority of respondents were in their 3rd or 4th year of study (@tbl-year-dist). The course's prerequisite of general statistics, which is a two-course sequence typically completed by students in their second year, would make it difficult to take this course earlier than 3rd year.

```{r tbl-year-dist}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: \"What year are you?\"

all_data |> 
  mutate(
    what_year_are_you = case_when(
      what_year_are_you == "PEY" ~ "4th", # Change PEY to 4th year
      what_year_are_you == "5th" ~ "5th or over",
      what_year_are_you == "6th" ~ "5th or over",
      TRUE ~ what_year_are_you
    )
  ) |> 
  count(what_year_are_you) |>
  pivot_wider(names_from = what_year_are_you, 
              values_from = n, 
              values_fill = list(n = 0)) |> 
  kable()
```

Respondents had a varied self-perception of their coding and writing abilities (@fig-selfperception). Most respondents believed it is important to be good at writing, but many are either indifferent or do not like to write (@fig-selfperception-1). Respondents also do not find writing to be particularly easy, which could be associated with the reported relative antipathy toward writing. Most respondents were at least somewhat confident in their own writing abilities, but a substantial contingent felt otherwise. Although few respondents felt that they were confident in their writing ability, more felt that they were able to catch their mistakes, which could indicate a disconnect between how respondents perceive their work and how the work was evaluated.

```{r fig-selfperception}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: Self-perception of coding and writing abilities
#| fig-subcap: ["\"Please rate how much each statement describes you, on a scale from 'This is very different to me' to 'This is a lot like me'\"", "\"Please rate how much each statement describes you, on a scale from 'This is very different to me' to 'This is a lot like me'\""]

# Writing
writing_perceptions_long <- 
  all_data |>
  select(
    writing_is_easy_for_me, 
    i_like_to_write, 
    i_believe_it_is_important_to_be_a_good_writer,
    when_i_edit_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_writing,
    i_am_confident_in_my_overall_writing_ability
  ) |> 
  pivot_longer(cols = everything(), names_to = "statements", values_to = "response") |>
  mutate(
    statements = case_when(
      statements == "writing_is_easy_for_me" ~ "\"Writing is easy for me.\"",
      statements == "i_like_to_write" ~ "\"I like to write.\"",
      statements == "i_believe_it_is_important_to_be_a_good_writer" ~ "\"I believe it is important\nto be a good writer.\"",
      statements == "when_i_edit_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I edit it is easy for\nme to catch my mistakes.\"",
      statements == "i_feel_confident_sharing_my_writing" ~ "\"I feel confident\nsharing my writing.\"",
      statements == "i_am_confident_in_my_overall_writing_ability" ~ "\"I am confident in my\noverall writing ability.\"",
      TRUE ~ statements
    ),
    response = case_when(
      response == "\"This is a lot like me\"" ~ "A lot like me",
      response == "\"This is very different to me\"" ~ "Very different from me",
      response == "\"This somewhat describes me\"" ~ "Somewhat describes me",
      TRUE ~ response
    )
  )

writing_perceptions_long |>
  ggplot(aes(x = response, fill = response)) +
  geom_bar(position = "stack", stat = "count") +
  facet_wrap(vars(statements)) +
  labs(
    x = "Response",
    y = "Number of repondents",
    fill = "Response:"
  ) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    legend.position = "bottom"
  )

# Coding
coding_perceptions_long <- 
  all_data |>
  select(
    coding_is_easy_for_me,
    i_like_to_code,
    i_believe_it_is_important_to_be_a_good_coder,
    when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes,
    i_feel_confident_sharing_my_code,
    i_am_confident_in_my_overall_coding_ability
  ) |> 
  pivot_longer(cols = everything(), names_to = "statements", values_to = "response") |>
  mutate(
    statements = case_when(
      statements == "coding_is_easy_for_me" ~ "\"Coding is easy for me\"",
      statements == "i_like_to_code" ~ "\"I like to code\"",
      statements == "i_believe_it_is_important_to_be_a_good_coder" ~ "\"I believe it is important\n to be a good coder\"",
      statements == "when_i_check_my_code_it_is_easy_for_me_to_catch_my_mistakes" ~ "\"When I check my code, it is\n easy for me to catch my\nmistakes\"",
      statements == "i_feel_confident_sharing_my_code" ~ "\"I feel confident\n sharing my code\"",
      statements == "i_am_confident_in_my_overall_coding_ability" ~ "\"I am confident in\n my overall coding ability\"",
      TRUE ~ statements
    ),
    response = case_when(
      response == "\"This is a lot like me\"" ~ "A lot like me",
      response == "\"This is very different to me\"" ~ "Very different from me",
      response == "\"This somewhat describes me\"" ~ "Somewhat describes me",
      TRUE ~ response
    )
  )

coding_perceptions_long |>
  ggplot(aes(x = response, fill = response)) +
  geom_bar(position = "stack", stat = "count") +
  facet_wrap(vars(statements)) +
  labs(
    x = "Response",
    y = "Number of repondents",
    fill = "Response:"
  ) +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    legend.position = "bottom"
  )
```

Respondent self-perceptions regarding coding proficiency and importance varied (@fig-selfperception-2). There was strong consensus on the perceived importance of coding skills, with a majority of respondents strongly affirming this belief. However, respondents' self-assessed coding abilities and enjoyment are more heterogeneous, with a substantial proportion reporting moderate rather than high levels of ease and enjoyment in coding tasks. These findings suggest a complicated relationship between respondents' recognition of coding's importance and their personal experience with coding.

Overall there was a moderate level of confidence among respondents in their overall coding ability, willingness to share code, and capacity to identify errors (@fig-selfperception-2). Notably, respondents express slightly higher confidence in detecting their own coding mistakes compared to general coding ability or code sharing. These patterns suggest that while respondents have developed some coding self-efficacy, there is still considerable potential for enhancing their perceived competence and comfort across various coding-related activities.

The majority of respondents were at least "somewhat familiar" with generative AI such as ChatGPT (@tbl-ai-familiarity-usage-1). Though respondents' self-perceptions around writing and coding are varied, there was strong consensus that the use of generative AI tools is appropriate within an academic setting (@tbl-ai-familiarity-usage-2). Most respondents who responded "It depends" generally found artificial intelligence tools to be appropriate, though with certain guidelines and rules governing their use.

```{r tbl-ai-familiarity-usage}
#| message: false
#| echo: false
#| warning: false
#| tbl-cap: Familiarity with, and appropriateness of, generative AI
#| tbl-subcap: ["'How familiar are you with using generative AI tools such as OpenAI's ChatGPT or equivalents?'","'To what extent do you think using generative AI tools such as ChatGPT by OpenAI (or equivalents) is ethical and appropriate for schoolwork?'"]
